---
title: "BT-474 Breast Cancer Cell Line Analysis"
author: "Chiraag Gohel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BT-474 Breast Cancer Cell Line Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Load required libraries
library(massSight)
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(purrr)
library(janitor)
library(ggrepel)
library(cowplot)
library(RColorBrewer)
library(cluster)
```

## Introduction

This analysis examines two independent LC-MS datasets from BT-474 breast cancer cell lines treated with fatty acid synthesis inhibitors. We'll use massSight to integrate these datasets and perform comprehensive downstream analysis including differential expression, pathway analysis, and multivariate statistical modeling.

The data can be downloaded from [here](https://gwu.box.com/s/0fhjdkd80bszoxjcjxyi7647osld3luc).

## Data Import and Processing

First, we'll import the data from the original BT-474 files and create massSight objects:

```{r, echo = F}
# Define paths to BT-474 data directories
bt_directories <- list.dirs("/Users/chiraag/Box/massSight/data/bt-474", recursive = FALSE) |>
  list.dirs(recursive = FALSE) |>
  list.dirs(recursive = FALSE)
```

```{r, eval = F}
bt_directories <- list.dirs("/path/to/bt-474", recursive = FALSE) |>
  list.dirs(recursive = FALSE) |>
  list.dirs(recursive = FALSE)
```

```{r}
# Create named list using directory structure
bt_data <- bt_directories |>
  purrr::map(\(path) {
    # Extract sample name and polarity from path
    parts <- str_split(path, "/")[[1]]
    sample_id <- parts[length(parts) - 2] # e.g. "st003147"
    polarity <- str_extract(parts[length(parts) - 1], "(?<=-)\\w+") # extracts "neg" from "pcpfm-neg"
    name <- paste(sample_id, polarity, sep = "_") # creates "st003147_neg"

    raw_table <- read_delim(paste0(path, "/output/preferred_no_blanks_Feature_table.tsv")) |>
      clean_names()
    normalized_table <- read_delim(paste0(path, "/output/normalized_Feature_table.tsv")) |>
      clean_names()
    # Get the asari project directory by finding the directory matching the pattern
    asari_dir <- list.files(path, pattern = "asari.*project.*", full.names = TRUE)
    annotations <- read_delim(file.path(asari_dir, "Feature_annotation.tsv")) |>
      clean_names()

    ms_data <- tibble(
        id_number = normalized_table$id_number,
        mz = raw_table$mz,
        rt = raw_table$rtime/60,
        intensity = normalized_table$peak_area
    ) |>
        left_join(
            annotations |>
                select(peak_id_number, emp_cpd_ion_relation, name_1st_guess),
            by = c("id_number" = "peak_id_number")
        )
    
    # Filter for specific adducts based on polarity
    if (polarity == "neg") {
        ms_data <- ms_data |>
            filter(emp_cpd_ion_relation == "M0,M-H-")
    } else {
        ms_data <- ms_data |>
            filter(emp_cpd_ion_relation == "M0,M+H+")
    }

    list(
      name = name,
      ms_data = ms_data,
      raw_table = raw_table,
      normalized_table = normalized_table,
      annotations = annotations
    )
  }) |>
  set_names(map_chr(bt_directories, \(path) {
    parts <- str_split(path, "/")[[1]]
    sample_id <- parts[length(parts) - 2]
    polarity <- str_extract(parts[length(parts) - 1], "(?<=-)\\w+")
    paste(sample_id, polarity, sep = "_")
  }))
```

## Creating massSight Objects

Transform the data into massSight objects for alignment:

```{r create-ms-objects}
ms_objects <- bt_data |>
    map(\(x) {
        create_ms_obj(
            x$ms_data,
            name = x$name,
            id_name = "id_number",
            mz_name = "mz",
            rt_name = "rt",
            int_name = "intensity",
            metab_name = "name_1st_guess"
        )
    })
```

## Aligning Datasets with massSight

Use the `mass_combine` function to align metabolites across datasets:

```{r combine-datasets}
# Negative mode datasets
neg_out <- mass_combine(ms_objects$st003147_neg,
    ms_objects$st003148_neg,
    log = NULL
)

# Positive mode datasets
pos_out <- mass_combine(ms_objects$st003147_pos,
    ms_objects$st003148_pos,
    log = NULL
)

# Visualize alignment results
final_plots(neg_out)
final_plots(pos_out)
```

## Loading and Processing Sample Metadata

```{r process-metadata}
# Load metadata
metadata_47_pos <- jsonlite::fromJSON("/Users/chiraag/Box/massSight/data/bt-474/st003147/positive.json")
metadata_47_neg <- jsonlite::fromJSON("/Users/chiraag/Box/massSight/data/bt-474/st003147/negative.json")
metadata_48_pos <- jsonlite::fromJSON("/Users/chiraag/Box/massSight/data/bt-474/st003148/positive.json")
metadata_48_neg <- jsonlite::fromJSON("/Users/chiraag/Box/massSight/data/bt-474/st003148/negative.json")

# Process metadata
metadata_47_neg <- bind_cols(
    "id" = metadata_47_neg$SUBJECT_SAMPLE_FACTORS$`Sample ID`,
    metadata_47_neg$SUBJECT_SAMPLE_FACTORS$Factors
) |>
    as_tibble() |>
    mutate(study = "47") |>
    clean_names()

metadata_48_neg <- bind_cols(
    "id" = metadata_48_neg$SUBJECT_SAMPLE_FACTORS$`Sample ID`,
    metadata_48_neg$SUBJECT_SAMPLE_FACTORS$Factors
) |>
    as_tibble() |>
    mutate(study = "48") |>
    clean_names()

metadata_47_pos <- bind_cols(
    "id" = metadata_47_pos$SUBJECT_SAMPLE_FACTORS$`Sample ID`,
    metadata_47_pos$SUBJECT_SAMPLE_FACTORS$Factors
) |>
    as_tibble() |>
    mutate(study = "47") |>
    clean_names()

metadata_48_pos <- bind_cols(
    "id" = metadata_48_pos$SUBJECT_SAMPLE_FACTORS$`Sample ID`,
    metadata_48_pos$SUBJECT_SAMPLE_FACTORS$Factors
) |>
    as_tibble() |>
    mutate(study = "48") |>
    clean_names()

# Combine metadata
pos_metadata <- bind_rows(metadata_47_pos, metadata_48_pos) |>
    mutate(drug_treatment = ifelse(drug_treatment == "NA" | drug_treatment == "Vehicle", "control", drug_treatment)) |>
    mutate(drug_treatment = factor(drug_treatment, levels = c("control", "1 uM Fasnall","1 uM GSK2194069")))

neg_metadata <- bind_rows(metadata_47_neg, metadata_48_neg) |>
    mutate(drug_treatment = ifelse(drug_treatment == "NA" | drug_treatment == "Vehicle", "control", drug_treatment)) |>
    mutate(drug_treatment = factor(drug_treatment, levels = c("control", "1 uM Fasnall","1 uM GSK2194069")))
```

## Extracting and Integrating Feature Data

```{r extract-features}
# Extract normalized features instead of raw
st003147_neg <- bt_data$st003147_neg$normalized_table |>
    select(id_number, contains("m22")) |>
    rename("Compound_ID_st003147_neg" = id_number)

st003148_neg <- bt_data$st003148_neg$normalized_table |>
    select(id_number, contains("bt474")) |>
    rename("Compound_ID_st003148_neg" = id_number)

# Extract positive mode features
st003147_pos <- bt_data$st003147_pos$normalized_table |>
    select(id_number, contains("m22")) |>
    rename("Compound_ID_st003147_pos" = id_number)

st003148_pos <- bt_data$st003148_pos$normalized_table |>
    select(id_number, contains("bt474")) |>
    rename("Compound_ID_st003148_pos" = id_number)

# Get feature mapping from mass_combine results
feature_mapping <- neg_out |>
    get_unique_matches() |>
    select(Compound_ID_st003147_neg, Compound_ID_st003148_neg, rep_Compound_ID, 
           Metabolite_st003147_neg, Metabolite_st003148_neg)

# Get unique matches from negative mode dataset
full_neg <- neg_out |>
    get_unique_matches() |>
    left_join(st003147_neg, by = "Compound_ID_st003147_neg") |>
    left_join(st003148_neg, by = "Compound_ID_st003148_neg") |>
    select(contains("Compound_ID"), contains("RT"), contains("MZ"), contains("Metabolite"), contains("m22"), contains("bt474")) |>
    pivot_longer(
        cols = c(contains("m22"), contains("bt474")),
        names_to = "sample",
        values_to = "intensity"
    ) |>
    left_join(
        neg_metadata |>
            mutate(id = tolower(gsub(
                "-", "_",
                ifelse(grepl("^[0-9]", id),
                    paste0("x", id),
                    id
                )
            ))),
        by = c("sample" = "id")
    )

# Filter for high-prevalence features
high_prevalence_neg <- full_neg |>
    group_by(study, rep_Compound_ID) |>
    summarise(
        n_samples = n(),
        n_detected = sum(!is.na(intensity) & intensity > 0),
        prevalence = n_detected / n_samples,
        .groups = "drop"
    ) |>
    group_by(rep_Compound_ID) |>
    filter(n() == 2,  # Ensure compound appears in both studies
           all(prevalence > 0.8)) |>  # High prevalence in both
    select(study, rep_Compound_ID) |>
    left_join(full_neg, by = c("study", "rep_Compound_ID")) |>
    mutate(drug_treatment = factor(drug_treatment, levels = c("control", "1 uM Fasnall","1 uM GSK2194069")))

# Get unique matches from positive mode dataset (similar process)
# Get unique matches from positive mode dataset
full_pos <- pos_out |>
    get_unique_matches() |>
    left_join(st003147_pos, by = "Compound_ID_st003147_pos") |>
    left_join(st003148_pos, by = "Compound_ID_st003148_pos") |>
    select(contains("Compound_ID"), contains("RT"), contains("MZ"), contains("Metabolite"), contains("m22"), contains("bt474")) |>
    pivot_longer(
        cols = c(contains("m22"), contains("bt474")),
        names_to = "sample", 
        values_to = "intensity"
    ) |>
    left_join(
        pos_metadata |>
            mutate(id = tolower(gsub(
                "-", "_",
                ifelse(grepl("^[0-9]", id),
                    paste0("x", id),
                    id
                )
            ))),
        by = c("sample" = "id")
    )

# Filter for high-prevalence features
high_prevalence_pos <- full_pos |>
    group_by(study, rep_Compound_ID) |>
    summarise(
        n_samples = n(),
        n_detected = sum(!is.na(intensity) & intensity > 0),
        prevalence = n_detected / n_samples,
        .groups = "drop"
    ) |>
    group_by(rep_Compound_ID) |>
    filter(n() == 2,  # Ensure compound appears in both studies
           all(prevalence > 0.8)) |>  # High prevalence in both
    select(study, rep_Compound_ID) |>
    left_join(full_pos, by = c("study", "rep_Compound_ID")) |>
    mutate(drug_treatment = factor(drug_treatment, levels = c("control", "1 uM Fasnall","1 uM GSK2194069")))

```

## Differential Expression Analysis

```{r differential-analysis}
# Negative mode linear modeling
lm_neg <- high_prevalence_neg |>
    mutate(drug_treatment = factor(drug_treatment, levels = c("control", "1 uM Fasnall","1 uM GSK2194069")),
           study = factor(study, levels = c("47", "48"))) |>
    group_by(rep_Compound_ID) |>
    nest() |>
    mutate(
        model = map(data, ~lm(log2(intensity + 1) ~ drug_treatment + study, data = .)),
        tidied = map(model, broom::tidy)
    ) |>
    unnest(tidied) |>
    select(rep_Compound_ID, term, estimate, std.error, statistic, p.value) |>
    mutate(
        fdr = p.adjust(p.value, method = "BH")
    )

# Filter for GSK2194069 results
gsk_results_neg <- lm_neg |>
    filter(term == "drug_treatment1 uM GSK2194069") |>
    left_join(
        neg_out |>
            get_unique_matches() |>
            select(
                rep_Compound_ID,
                contains("Compound_ID"),
                contains("Metabolite"),
                contains("MZ"),
                contains("RT")
            ),
        by = "rep_Compound_ID"
    )

# Filter for Fasnall results
fas_results_neg <- lm_neg |>
    filter(term == "drug_treatment1 uM Fasnall") |>
    left_join(
        neg_out |>
            get_unique_matches() |>
            select(
                rep_Compound_ID,
                contains("Compound_ID"),
                contains("Metabolite"),
                contains("MZ"),
                contains("RT")
            ),
        by = "rep_Compound_ID"
    )

# Summary of differential metabolites
sig_threshold <- 0.05
fc_threshold <- 1

gsk_sig_metabolites <- gsk_results_neg |> 
    filter(fdr < sig_threshold & abs(estimate) > fc_threshold)
fas_sig_metabolites <- fas_results_neg |> 
    filter(fdr < sig_threshold & abs(estimate) > fc_threshold)

cat("GSK2194069 significant metabolites:", nrow(gsk_sig_metabolites), "\n")
cat("Fasnall significant metabolites:", nrow(fas_sig_metabolites), "\n")

# Create similar analysis for positive mode data
# Fit linear models for positive mode data
lm_pos <- full_pos |>
    mutate(drug_treatment = factor(drug_treatment, levels = c("control", "1 uM Fasnall","1 uM GSK2194069"))) |>
    group_by(rep_Compound_ID) |>
    nest() |>
    mutate(
        model = map(data, \(d) lm(log2(intensity + 1) ~ drug_treatment + study, data = d)),
        tidied = map(model, broom::tidy)
    ) |>
    unnest(tidied) |>
    select(rep_Compound_ID, term, estimate, std.error, statistic, p.value) |>
    mutate(
        fdr = p.adjust(p.value, method = "BH")
    )

# Filter for GSK2194069 results
gsk_results_pos <- lm_pos |>
    filter(term == "drug_treatment1 uM GSK2194069") |>
    left_join(
        pos_out |>
            get_unique_matches() |>
            select(
                rep_Compound_ID,
                contains("Metabolite"),
                contains("MZ"),
                contains("RT")
            ),
        by = "rep_Compound_ID"
    )

# Filter for Fasnall results
fas_results_pos <- lm_pos |>
    filter(term == "drug_treatment1 uM Fasnall") |>
    left_join(
        pos_out |>
            get_unique_matches() |>
            select(
                rep_Compound_ID,
                contains("Metabolite"),
                contains("MZ"),
                contains("RT")
            ),
        by = "rep_Compound_ID"
    )

# Summary of differential metabolites for positive mode
gsk_sig_metabolites_pos <- gsk_results_pos |> 
    filter(fdr < sig_threshold & abs(estimate) > fc_threshold)
fas_sig_metabolites_pos <- fas_results_pos |> 
    filter(fdr < sig_threshold & abs(estimate) > fc_threshold)

cat("GSK2194069 significant metabolites (positive mode):", nrow(gsk_sig_metabolites_pos), "\n")
cat("Fasnall significant metabolites (positive mode):", nrow(fas_sig_metabolites_pos), "\n")

```

## Volcano Plots for Visualization

```{r volcano-plots}
# Negative mode GSK2194069 plot
neg_gsk_plot <- gsk_results_neg |>
    mutate(
        significant = abs(estimate) > fc_threshold & fdr < sig_threshold,
        label = ifelse(abs(estimate) > 3 & fdr < sig_threshold,
            paste0(
                "ST003147: ",
                ifelse(is.na(Metabolite_st003147_neg), 
                    paste0("m/z:", round(MZ_st003147_neg, 4), " RT:", round(RT_st003147_neg, 2)),
                    Metabolite_st003147_neg),
                "\n",
                "ST003148: ",
                ifelse(is.na(Metabolite_st003148_neg),
                    paste0("m/z:", round(MZ_st003148_neg, 4), " RT:", round(RT_st003148_neg, 2)),
                    Metabolite_st003148_neg)
            ),
            "")
    ) |>
    ggplot(aes(x = estimate, y = -log10(fdr))) +
    geom_hline(yintercept = -log10(sig_threshold), linetype = "dashed", size = 0.3) +
    geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed", size = 0.3) +
    geom_point(aes(color = significant), size = 1.2, alpha = 0.7) +
    geom_label_repel(
        aes(label = label),
        max.overlaps = 20,
        size = 2.5,
        force = 10
    ) +
    scale_color_manual(values = c("grey", "red")) +
    labs(
        x = "Log2 Fold Change (GSK2194069 vs Control)",
        y = "-log10(FDR)",
        title = "Negative Mode - GSK2194069"
    ) +
    theme_minimal() +
    theme(legend.position = "none")

# Negative mode Fasnall plot
neg_fas_plot <- fas_results_neg |>
    mutate(
        significant = abs(estimate) > fc_threshold & fdr < sig_threshold,
        label = ifelse(abs(estimate) > 3 & fdr < sig_threshold,
            paste0(
                "ST003147: ",
                ifelse(is.na(Metabolite_st003147_neg), 
                    paste0("m/z:", round(MZ_st003147_neg, 4), " RT:", round(RT_st003147_neg, 2)),
                    Metabolite_st003147_neg),
                "\n",
                "ST003148: ",
                ifelse(is.na(Metabolite_st003148_neg),
                    paste0("m/z:", round(MZ_st003148_neg, 4), " RT:", round(RT_st003148_neg, 2)),
                    Metabolite_st003148_neg)
            ),
            "")
    ) |>
    ggplot(aes(x = estimate, y = -log10(fdr))) +
    geom_hline(yintercept = -log10(sig_threshold), linetype = "dashed", size = 0.3) +
    geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed", size = 0.3) +
    geom_point(aes(color = significant), size = 1.2, alpha = 0.7) +
    geom_label_repel(
        aes(label = label),
        max.overlaps = 20,
        size = 2.5,
        force = 10
    ) +
    scale_color_manual(values = c("grey", "red")) +
    labs(
        x = "Log2 Fold Change (Fasnall vs Control)",
        y = "-log10(FDR)", 
        title = "Negative Mode - Fasnall"
    ) +
    theme_minimal() +
    theme(legend.position = "none")

# Positive mode GSK2194069 plot
pos_gsk_plot <- gsk_results_pos |>
    mutate(
        significant = abs(estimate) > fc_threshold & fdr < sig_threshold,
        label = ifelse(abs(estimate) > 3 & fdr < sig_threshold,
            paste0(
                "ST003147: ",
                ifelse(is.na(Metabolite_st003147_pos), 
                    paste0("m/z:", round(MZ_st003147_pos, 4), " RT:", round(RT_st003147_pos, 2)),
                    Metabolite_st003147_pos),
                "\n",
                "ST003148: ",
                ifelse(is.na(Metabolite_st003148_pos),
                    paste0("m/z:", round(MZ_st003148_pos, 4), " RT:", round(RT_st003148_pos, 2)),
                    Metabolite_st003148_pos)
            ),
            "")
    ) |>
    ggplot(aes(x = estimate, y = -log10(fdr))) +
    geom_hline(yintercept = -log10(sig_threshold), linetype = "dashed", size = 0.3) +
    geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed", size = 0.3) +
    geom_point(aes(color = significant), size = 1.2, alpha = 0.7) +
    geom_label_repel(
        aes(label = label),
        max.overlaps = 20,
        size = 2.5,
        force = 10
    ) +
    scale_color_manual(values = c("grey", "red")) +
    labs(
        x = "Log2 Fold Change (GSK2194069 vs Control)",
        y = "-log10(FDR)",
        title = "Positive Mode - GSK2194069"
    ) +
    theme_minimal() +
    theme(legend.position = "none")

# Positive mode Fasnall plot
pos_fas_plot <- fas_results_pos |>
    mutate(
        significant = abs(estimate) > fc_threshold & fdr < sig_threshold,
        label = ifelse(abs(estimate) > 3 & fdr < sig_threshold,
            paste0(
                "ST003147: ",
                ifelse(is.na(Metabolite_st003147_pos), 
                    paste0("m/z:", round(MZ_st003147_pos, 4), " RT:", round(RT_st003147_pos, 2)),
                    Metabolite_st003147_pos),
                "\n",
                "ST003148: ",
                ifelse(is.na(Metabolite_st003148_pos),
                    paste0("m/z:", round(MZ_st003148_pos, 4), " RT:", round(RT_st003148_pos, 2)),
                    Metabolite_st003148_pos)
            ),
            "")
    ) |>
    ggplot(aes(x = estimate, y = -log10(fdr))) +
    geom_hline(yintercept = -log10(sig_threshold), linetype = "dashed", size = 0.3) +
    geom_vline(xintercept = c(-fc_threshold, fc_threshold), linetype = "dashed", size = 0.3) +
    geom_point(aes(color = significant), size = 1.2, alpha = 0.7) +
    geom_label_repel(
        aes(label = label),
        max.overlaps = 20,
        size = 2.5,
        force = 10
    ) +
    scale_color_manual(values = c("grey", "red")) +
    labs(
        x = "Log2 Fold Change (Fasnall vs Control)",
        y = "-log10(FDR)",
        title = "Positive Mode - Fasnall"
    ) +
    theme_minimal() +
    theme(legend.position = "none")

# Combine all plots
plot_grid(neg_gsk_plot, neg_fas_plot, pos_gsk_plot, pos_fas_plot, ncol = 2)
```

## Enhanced Visualization with Heatmaps

```{r heatmaps}
# Extract top differential metabolites for heatmap
top_metabolites <- rbind(
    gsk_results_neg |> filter(fdr < 0.01, abs(estimate) > 1.5) |> head(30),
    fas_results_neg |> filter(fdr < 0.01, abs(estimate) > 1.5) |> head(30)
) |>
    distinct(rep_Compound_ID, .keep_all = TRUE) 

# Create expression matrix for heatmap
heatmap_data <- high_prevalence_neg |>
    filter(rep_Compound_ID %in% top_metabolites$rep_Compound_ID) |>
    group_by(rep_Compound_ID, drug_treatment, study) |>
    summarise(avg_intensity = mean(log2(intensity + 1), na.rm = TRUE), .groups = "drop") |>
    unite("group", c(drug_treatment, study), remove = FALSE) |>
    pivot_wider(id_cols = rep_Compound_ID, names_from = group, values_from = avg_intensity) |>
    as.data.frame()

row_names <- top_metabolites |>
    select(rep_Compound_ID, Metabolite_st003147_neg) |>
    mutate(name = ifelse(is.na(Metabolite_st003147_neg), 
                         paste0("Metabolite ", rep_Compound_ID), 
                         Metabolite_st003147_neg)) |>
    pull(name)

# Convert to matrix for heatmap
mat <- as.matrix(heatmap_data[,-1])
rownames(mat) <- row_names

# Create annotation for columns
treatment_colors <- c("control" = "grey80", "1 uM Fasnall" = "skyblue", "1 uM GSK2194069" = "orange")
study_colors <- c("47" = "purple", "48" = "green")

col_anno <- data.frame(
    Treatment = gsub("_.*$", "", colnames(mat)),
    Study = gsub("^.*_", "", colnames(mat))
)


column_ha <- ComplexHeatmap::HeatmapAnnotation(
    Treatment = col_anno$Treatment,
    Study = col_anno$Study,
    col = list(
        Treatment = treatment_colors,
        Study = study_colors
    )
)

# Draw heatmap
hm <- ComplexHeatmap::Heatmap(
    mat,
    name = "log2 Intensity",
    column_title = "Top Differential Metabolites",
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    show_column_names = FALSE,
    top_annotation = column_ha,
    row_names_gp = grid::gpar(fontsize = 8)
)

# Display heatmap
hm
```

## Multivariate Analysis

```{r multivariate-analysis}
# Prepare data for PLS-DA
plsda_data <- high_prevalence_neg |>
    select(rep_Compound_ID, sample, intensity, drug_treatment, study) |>
    pivot_wider(
        id_cols = c(sample, drug_treatment, study),
        names_from = rep_Compound_ID,
        values_from = intensity,
        values_fill = 0
    )

# Extract X (metabolites) and Y (treatment groups) matrices
X <- as.matrix(log2(plsda_data[, 4:ncol(plsda_data)] + 1))
Y <- plsda_data$drug_treatment

# Perform PLS-DA
plsda_result <- mixOmics::plsda(X, Y, ncomp = 5)

# Plot results
mixOmics::plotIndiv(plsda_result, 
          group = Y,
          legend = TRUE,
          title = "PLS-DA of BT-474 Treatments",
          ellipse = TRUE,
          ind.names = TRUE)

# Variable importance
vip_scores <- mixOmics::vip(plsda_result)
top_vip <- data.frame(
    feature = colnames(X),
    VIP = vip_scores[,1]
) |>
    arrange(desc(VIP)) |>
    head(20)

# Plot VIP scores
ggplot(top_vip, aes(x = reorder(feature, VIP), y = VIP)) +
    geom_bar(stat = "identity", fill = "skyblue") +
    coord_flip() +
    labs(x = "Metabolite Feature", y = "VIP Score", 
         title = "Top 20 Features by Variable Importance")
```

## Exploring Treatment-Specific Effects

```{r treatment-comparison}
# Compare GSK vs Fasnall effects
comparison_df <- inner_join(
    gsk_results_neg |> select(rep_Compound_ID, gsk_estimate = estimate, gsk_fdr = fdr),
    fas_results_neg |> select(rep_Compound_ID, fas_estimate = estimate, fas_fdr = fdr),
    by = "rep_Compound_ID"
) |>
    mutate(
        gsk_significant = gsk_fdr < sig_threshold & abs(gsk_estimate) > fc_threshold,
        fas_significant = fas_fdr < sig_threshold & abs(fas_estimate) > fc_threshold,
        both_significant = gsk_significant & fas_significant,
        gsk_only = gsk_significant & !fas_significant,
        fas_only = fas_significant & !gsk_significant,
        effect_category = case_when(
            both_significant & sign(gsk_estimate) == sign(fas_estimate) ~ "Shared Effect",
            both_significant & sign(gsk_estimate) != sign(fas_estimate) ~ "Opposite Effect",
            gsk_only ~ "GSK-Specific",
            fas_only ~ "Fasnall-Specific",
            TRUE ~ "Non-Significant"
        )
    )

# Summarize comparison results
comparison_summary <- comparison_df |>
    group_by(effect_category) |>
    summarise(count = n()) |>
    arrange(desc(count))

# Scatter plot comparing treatments
comparison_plot <- ggplot(comparison_df, aes(x = gsk_estimate, y = fas_estimate, color = effect_category)) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", size = 0.3) +
    geom_vline(xintercept = 0, linetype = "dashed", size = 0.3) +
    scale_color_brewer(palette = "Set1") +
    labs(
        x = "Log2 Fold Change (GSK2194069)",
        y = "Log2 Fold Change (Fasnall)",
        title = "Comparison of Treatment Effects",
        color = "Effect Category"
    ) +
    theme_minimal()

# Display comparison plot
comparison_plot
```

