---
title: "Profiling software comparison using iHMP data"
author: "Bahar Sayoldin, bahar.sayoldin@gmail.com; Chiraag Gohel, chiraaggohel@gwu.edu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling software comparison using iHMP data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

`massSight` can be installed via `devtools`:

```{r, eval = F}
install.packages('devtools')
devtools::install_github("omicsEye/massSight", force = TRUE)
```

We can then load the necessary libraries.

```{r}
library(ggplot2)
library(massSight)
library(knitr)
```

## Loading iHMP data

For this tutorial, we will be analyzing two datasets from the NIH human microbiome project. The first dataset can be downloaded [here](https://gwu.box.com/v/HMP2metabolomics). The second dataset can be downloaded [here](https://gwu.box.com/s/hrgfgmlllvzvmzbzyn8tolygel9brkkz)

```{r}
loaded_data <- load_data(input='data/HMP2_metabolomics.xlsx', 
                         type = 'all', 
                         sheet = 1, 
                         id = 'Compound_ID')
loaded_data$feature_metadata$MZ <- as.numeric(loaded_data$feature_metadata$MZ)
loaded_data$feature_metadata$RT <- as.numeric(loaded_data$feature_metadata$RT)
feature_metadata2 <- loaded_data$feature_metadata[colnames(loaded_data$data),]
```

### Calculate the mean of intensity values across samples 

```{r}
loaded_data$data <- loaded_data$data |> t() |> data.frame()
hmp2_keep <- filter_intensities(data = loaded_data$data, prevalence = .5)
loaded_data$data <- loaded_data$data[hmp2_keep, ]
feature_metadata2 <- feature_metadata2[hmp2_keep, ]
feature_metadata2$Intensity <- rowMeans(loaded_data$data, na.rm=T)
ref_input <- feature_metadata2[(!is.na(feature_metadata2$MZ)) & (!is.na(feature_metadata2$RT)), ]
```

### create a `massSight` object as the first (ref) object for alignment

```{r}
ref <- create_ms_obj(df = ref_input, 
                     name = "iHMP", 
                     id_name = "Compound_ID", 
                     rt_name = "RT", 
                     mz_name = "MZ", 
                     int_name = "Intensity",
                     metab_name= "Metabolite")
```

### create a `massSight` object as a second (query) object for alignment

```{r}
C18_CD <- read.delim(
  'data/iHMP_C18-FFA_CompoundDixcoverer_Exported_All_Samples_Annotated.csv',
  sep = ',',
  header = TRUE,
  fill = FALSE,
  comment.char = "" ,
  check.names = FALSE
  #row.names = 1
)
```

We then check the column names to see what variable names should be used when converting the dataframe into a `massSight` object.

```{r}
colnames(C18_CD) |> head(10)
```

### calculate average intensity across all samples

In this dataset, sample intensity values begin at column 7 until the end of the dataframe.
The `load_data()` function used `omicsArt::numeric_dataframe()` to ensure that we converted the dataframe columns as numeric as the read dataframe has columns with various data types and to measure mean of rows of intensities we need to convert them to numeric.

```{r}
prev <- .5

row_nas <- C18_CD[,7:ncol(C18_CD)] |>
  is.na() |>
  rowSums()

keep <- (row_nas/ncol(C18_CD)) > prev

C18_CD_filter <- C18_CD[keep,]

C18_CD$Intensity <- rowMeans(omicsArt::numeric_dataframe(C18_CD[,7:dim(C18_CD)[2]]) , na.rm=T)
C18_CD$row_id <- rownames(C18_CD)
```

Create an object for C18_CD which includes FFA metabolites processed with Compound Discovery version

```{r}
ms_C18_CD <-
  create_ms_obj(
    df = C18_CD,
    name = "C18_CD",
    id_name = "row_id",
    rt_name = "RT [min]",
    mz_name = "m/z",
    int_name = "Intensity",
    metab_name= "Name"
  )
```

## check few first rows of the dataframe

```{r}
ms_C18_CD |>
  raw_df() |>
  head() |>
  kableExtra::kbl(format = "simple")
```

## run auto-align function with defualts 

```{r}
aligned <- auto_align(ms1 = ref, 
                      ms2 = ms_C18_CD,
                      rt_lower = -.25,
                      rt_upper = .25,
                      mz_lower = -5,
                      mz_upper = 5,
                      iso_method = "dbscan")
```

## Visualization and save pdfs

```{r}
## Visualization 
final_smooth <- final_plots(aligned)

ggsave(filename='manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_all.png', plot=final_smooth, width = 7.2, height = 3.5, units = "in", dpi = 300)
```

## Use only C18-neg as reference 

```{r}

ref_input_C18 <- ref_input[ref_input$Method=="C18-neg",]
ref_C18 <- create_ms_obj(df = ref_input_C18, 
                         name = "iHMP_C18", 
                         id_name = "Compound_ID", 
                         rt_name = "RT", 
                         mz_name = "MZ", 
                         int_name = "Intensity",
                         metab_name= "Metabolite")
```

## run auto_align with dbscan

```{r}
aligned_c18 <- auto_align(ms1 = ref_C18, 
                          ms2 = ms_C18_CD, 
                          iso_method = "dbscan")
```

## Visualization and save pdfs

```{r}
final_smooth_ref_C18 <- final_plots(aligned_c18)
ggsave(filename='manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_C18.png', plot=final_smooth_ref_C18, width = 7.2, height = 3.5, units = "in", dpi = 300)
```

## run auto_align with default

```{r}
aligned_c18_default <- auto_align(ms1 = ref_C18, 
                                  ms2 = ms_C18_CD)

final_smooth_ref_C18_default <- final_plots(aligned_c18, iso_method = "manual")

ggsave(filename='manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_C18_default.png', plot=final_smooth_ref_C18_default, width = 7.2, height = 3.5, units = "in", dpi = 300)
```
