---
title: "Profiling software comparison using iHMP data"
author: "Bahar Sayoldin, bahar.sayoldin@gmail.com"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling software comparison using iHMP data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

install.packages('devtools')
library(devtools)
devtools::install_github("omicsEye/massSight", force = TRUE)

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load libraries including `massSight`
```{r}
library(ggplot2)
library(omicsArt)
library(massSight)
library(knitr)
```

## Set working directory
```{r setup, include=F, cache = F}
opts_knit$set(root.dir = "~/Library/CloudStorage/Box-Box/massSight")
```

## Loading two dataframes containing iHMP profiles.

```{r}
loaded_data <- massSight::load_data(input='data/iHMP_ProgenesisQI/HMP2_metabolomics.xlsx', type = 'all', sheet = 1, id = 'Compound_ID')
loaded_data$feature_metadata$MZ <- as.numeric(loaded_data$feature_metadata$MZ)
loaded_data$feature_metadata$RT <- as.numeric(loaded_data$feature_metadata$RT)
feature_metadata2 <-loaded_data$feature_metadata[colnames(loaded_data$data),]
```
## calculate mean of intensity across samples 

```{r}
feature_metadata2$Intensity <- colMeans(loaded_data$data, na.rm=T)
ref_input <- feature_metadata2[!is.na(feature_metadata2$MZ) & !is.na(feature_metadata2$RT), ]
```

## create a `massSight` object as the first (ref) object for alignment

```{r}
ref <- create_ms_obj(df = ref_input, 
                     name = "iHMP", 
                     id_name = "Compound_ID", 
                     rt_name = "RT", 
                     mz_name = "MZ", 
                     int_name = "Intensity",
                     metab_name= "Metabolite")
```

## create a `massSight` object as a second (query) object for alignment

```{r}
C18_CD <- read.delim(
  'data/iHMP_CompoundDiscoverer/C18neg_FFA method/iHMP_C18-FFA_CompoundDixcoverer_Exported_All_Samples_Annotated.csv',
  sep = ',',
  header = TRUE,
  fill = FALSE,
  comment.char = "" ,
  check.names = FALSE
  #row.names = 1
)
```

check the column names to see metabolite descriptors names to be used in objects creation

```{r}
colnames(C18_CD)
```

### calculate average intensity across all samples

In this dataframe samples start from col number 7 and goes to the end of the dataframe.
we used omicsArt::numeric_dataframe to make sure we convert the dataframe as numeric as the read dataframe has columns with various data types and to measure mean of rows of intensities we need to convert them to numeric.

```{r}
C18_CD$Intensity <- rowMeans(omicsArt::numeric_dataframe(C18_CD[,7:dim(C18_CD)[2]]) , na.rm=T)

C18_CD$row_id <- rownames(C18_CD)
```

Create an object for C18_CD which includes FFA metabolites processed with Compound Discovery version

## Creating `massSight` objects

```{r}
ms_C18_CD <-
  create_ms_obj(
    df = C18_CD,
    name = "C18_CD",
    id_name = "row_id",
    rt_name = "RT [min]",
    mz_name = "m/z",
    int_name = "Intensity",
    metab_name= "Name"
  )
```

## check few first rows of the dataframe

```{r}
ms_C18_CD |>
  raw_df() |>
  head() |>
  kableExtra::kbl(format = "simple")
```
## run aut-align function with defualts 

```{r}
aligned <- auto_align(ms1 = ref, 
                      ms2 = ms_C18_CD, 
                      iso_method = "dbscan")
```

## Visualization and save pdfs

```{r}
final_smooth <- final_plots(aligned)

ggsave(filename='manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_all.png', plot=final_smooth, width = 7.2, height = 3.5, units = "in", dpi = 300)
```

## Use only C18-neg as reference 

```{r}
ref_input_C18 <- ref_input[ref_input$Method=="C18-neg",]
ref_C18 <- create_ms_obj(df = ref_input_C18, 
                         name = "iHMP_C18", 
                         id_name = "Compound_ID", 
                         rt_name = "RT", 
                         mz_name = "MZ", 
                         int_name = "Intensity",
                         metab_name= "Metabolite")
```

## run auto_align with dbscan

```{r}
aligned_c18 <- auto_align(ms1 = ref_C18, 
                          ms2 = ms_C18_CD, 
                          iso_method = "dbscan")
```

## Visualization and save pdfs

```{r}
final_smooth_ref_C18 <- final_plots(aligned_c18)
ggsave(filename='manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_C18.png', plot=final_smooth_ref_C18, width = 7.2, height = 3.5, units = "in", dpi = 300)
```

## run auto_align with dafualt
```{r}
aligned_c18_default <- auto_align(ms1 = ref_C18, 
                                  ms2 = ms_C18_CD)

final_smooth_ref_C18_default <- final_plots(aligned_c18, iso_method = "manual")

ggsave(filename='manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_C18_default.png', plot=final_smooth_ref_C18_default, width = 7.2, height = 3.5, units = "in", dpi = 300)
```
