---
title: "Alignment Tutorial using iHMP Data"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
author:
- Chiraag Gohel, chiraaggohel@gwu.edu
- Bahar Sayoldin, bahar.sayoldin@gmail.com
- Ali Rahnavard, rahnavard@gwu.edu
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alignment Tutorial using iHMP Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

For this tutorial, we will be analyzing two datasets from the NIH human microbiome project that were processed using different techniques. The first dataset can be downloaded [here](https://gwu.box.com/v/HMP2metabolomics). The second dataset can be downloaded [here](https://gwu.box.com/s/hrgfgmlllvzvmzbzyn8tolygel9brkkz). 

First, `massSight` can be installed via `devtools`:

```{r, eval = F}
install.packages('devtools')
devtools::install_github("omicsEye/massSight", force = TRUE)
devtools::install_github("omicsEye/omicsArt", force = TRUE)
```

We can then load the necessary libraries. `omicsArt` provides helper functions for loading and visualizing data.

```{r, message = F}
library(massSight)
library(ggplot2)
library(omicsArt)
```

## Loading iHMP data

We can use the `load_data()` function to import LC-MS data in excel format from a variety of standard pre-processed formats.

```{r, cache = T, results = 'hide', message=FALSE, warning=FALSE}
loaded_data <-
  massSight::load_data(
    input = 'data/HMP2_metabolomics.xlsx',
    type = 'all',
    sheet = 1,
    id = 'Compound_ID'
  )
loaded_data$feature_metadata$MZ <-
  as.numeric(loaded_data$feature_metadata$MZ)
loaded_data$feature_metadata$RT <-
  as.numeric(loaded_data$feature_metadata$RT)
feature_metadata2 <-
  loaded_data$feature_metadata[colnames(loaded_data$data), ]
```

We can then use the `filter_intensities()` function to perform quality control and remove metabolites with low prevalence. 

```{r}
loaded_data$data <- loaded_data$data |> t() |> data.frame()
hmp2_keep <-
  filter_intensities(data = loaded_data$data, prevalence = .5)
loaded_data$data <- loaded_data$data[hmp2_keep,]
feature_metadata2 <- feature_metadata2[hmp2_keep,]
feature_metadata2$Intensity <- rowMeans(loaded_data$data, na.rm = T)
ref_input <-
  feature_metadata2[(!is.na(feature_metadata2$MZ)) &
                      (!is.na(feature_metadata2$RT)),]
```

### Create a `massSight` object for the first dataset

We now have everything we need to create a `massSight` object (`MSObject`). The object serves as a container that contains raw data, analyzed data, and other information regarding the experiment. For more information about the `MSObject`, check out its [documentation](https://omicseye.github.io/massSight/reference/MSObject-class.html).

```{r}
hmp2_ms <- create_ms_obj(
  df = ref_input,
  name = "iHMP",
  id_name = "Compound_ID",
  rt_name = "RT",
  mz_name = "MZ",
  int_name = "Intensity",
  metab_name = "Metabolite"
)
```


We can use the `raw_df()` function to access the stored data from the created object. Let's see what it looks like!

```{r}
hmp2_ms |>
  raw_df() |>
  head() |>
  kableExtra::kbl(format = "simple")
```

We process the second dataset similarly.

```{r}
C18_CD <- read.delim(
  'data/iHMP_C18-FFA_CompoundDixcoverer_Exported_All_Samples_Annotated.csv',
  sep = ',',
  header = TRUE,
  fill = FALSE,
  comment.char = "" ,
  check.names = FALSE
  #row.names = 1
)
```

We then check the column names to see what variable names should be used when converting the dataframe into a `massSight` object.

```{r}
colnames(C18_CD) |> head(10)
```

In this dataset, sample intensity values begin at column 7 until the end of the dataframe.
The `load_data()` function used `omicsArt::numeric_dataframe()` to ensure that we converted the dataframe columns as numeric as the read dataframe has columns with various data types and to measure mean of rows of intensities we need to convert them to numeric.

```{r}
c18_keep <- filter_intensities(data = C18_CD[, 7:ncol(C18_CD)],
                               prevalence = .5)
C18_CD_filter <- C18_CD[c18_keep, ]
C18_CD$Intensity <-
  rowMeans(omicsArt::numeric_dataframe(C18_CD[, 7:dim(C18_CD)[2]]),
           na.rm = T)
C18_CD$row_id <- rownames(C18_CD)
```

We then can create an object for C18_CD which includes FFA metabolites processed with Compound Discovery version

```{r}
ms_C18_CD <-
  create_ms_obj(
    df = C18_CD,
    name = "C18_CD",
    id_name = "row_id",
    rt_name = "RT [min]",
    mz_name = "m/z",
    int_name = "Intensity",
    metab_name= "Name"
  )
```

## Aligning Datasets

The `auto_combine()` function allows users to combine two datasets via the modeling of RT and m/z drift between the two experiments. For more information on the function, check out its [documentation](https://omicseye.github.io/massSight/reference/auto_combine.html)!

```{r, cache = T}
aligned <- auto_combine(
  ms1 = hmp2_ms,
  ms2 = ms_C18_CD,
  iso_method = "dbscan",
  eps = .15,
  smooth_method = "loess"
)
```

## Visualization

Visualization of alignment can be performed via the `final_plots()` function.

```{r}
final_plots(aligned)
```

We recommend the use of `ggsave()` from the package `ggplot2` for the saving of publication quality figures.

```{r, eval = F}
## Visualization
final_smooth <- final_plots(aligned)

ggsave(
  filename = 'manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_all.png',
  plot = final_smooth,
  width = 7.2,
  height = 3.5,
  units = "in",
  dpi = 300
)
```

## Using only C18-neg as a reference 

```{r}
ref_input_C18 <- ref_input[ref_input$Method == "C18-neg", ]
ref_C18 <- create_ms_obj(
  df = ref_input_C18,
  name = "iHMP_C18",
  id_name = "Compound_ID",
  rt_name = "RT",
  mz_name = "MZ",
  int_name = "Intensity",
  metab_name = "Metabolite"
)
```

## Run auto_combine with dbscan

```{r, cache = T}
aligned_c18 <- auto_combine(
  ms1 = ref_C18,
  ms2 = ms_C18_CD,
  iso_method = "dbscan",
  eps = .15,
  smooth_method = "loess"
)
```

## Visualization

```{r}
final_plots(aligned_c18)
```

```{r, eval = F}
final_smooth_ref_C18 <- final_plots(aligned_c18)
ggsave(
  filename = 'manuscript/figures/fig_iHMP_software_comp/final_smooth_ref_C18.png',
  plot = final_smooth_ref_C18,
  width = 7.2,
  height = 3.5,
  units = "in",
  dpi = 300
)
```
